%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Usage
%
\chapter{Synopsis}
\label{chapter:options}
\cutname{omake-options.html}

\Prog{omake}
    \oOpt{-k}
    \oOptArg{-j}{count}
    \oOpt{-n}
    \oOpt{-s} \oOpt{-S}
    \oOpt{-p}
    \oOpt{-P}
    \oOpt{-w}
    \oOpt{-t}
    \oOpt{-u}
    \oOpt{-U}
    \oOpt{-R}
    \oOpt{-{}-project}
    \oOpt{-{}-progress}
    \oOpt{-{}-print-status}
    \oOpt{-{}-print-exit}
    \oOpt{-{}-print-dependencies}
    \oOptArg{-{}-show-dependencies}{target}
    \oOpt{-{}-all-dependencies}
    \oOpt{-{}-verbose-dependencies}
    \oOpt{-{}-force-dotomake}
    \oOptArg{-{}-dotomake}{dir}
    \oOpt{-{}-flush-includes}
    \oOpt{-{}-configure}
    \oOpt{-{}-install}
    \oOpt{-{}-install-all}
    \oOpt{-{}-install-force}
    \oOpt{-{}-version}
    \oOpt{-{}-absname}
    \oOpt{-{}-divert-errors}
    \oOpt{-{}-divert-repeat}
    \oOpt{-{}-divert-only}
    \oArg{filename...}
    \oOpt{var-definition...}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Options
%
\section{General usage}

For Boolean options (for example, \verb+-s+, \verb+--progress+, etc.) the option can include a
prefix \verb+--no-+, which inverts the usual sense of the option.  For example, the option
\verb+--progress+ means ``print a progress bar,''  while the option \verb+--no-progress+ means
``do not print a progress bar.''

If multiple instances of an option are specified, the final option determines the behavior of OMake.
In the following command line, the final \verb+--no-S+ cancels the earlier \verb+-S+.

\begin{verbatim}
    % omake -S --progress --no-S
\end{verbatim}

\section{Output control}

\option{-s} \verb+-s+

Do not print commands as they are executed (be ``silent'').

\option{-S} \verb+-S+

Do not print commands as they are executed \emph{unless} they produce output and/or
fail.

\option{-w} \verb+-w+

Print directory information in \Prog{make} format as commands are executed.
This is mainly useful for editors that expect \Prog{make}-style
directory information for determining the location of errors.

\option{--progress} \verb+--progress+

Print a progress indicator.
This is normally used with the \Opt{-s} or \Opt{-S} options.

\option{--print-status} \verb+--print-status+

Print status lines (the \verb/+/ and \verb/-/ lines).

\option{--print-exit} \verb+--print-exit+

Print termination codes when commands complete.

\option{--divert-errors} \verb+--divert-errors+

``Divert'' command output.  The effect of this option to save the output of any commands so that it
can be printed at the end of the build.  If any rules/commands fail, then OMake will re-print the
output of the failed commands when it finishes the build.  Off by default.

\option{--divert-repeat} \verb+--divert-repeat+

When a rule finishes, print the output as a single block.  This is useful
in combination \verb+-j+ option (see Section~\ref{option:-j}), where the output of multiple subprocesses can be
garbled.  The diversion is printed as a single coherent unit.  Implies \verb+--divert-errors+.

\option{--divert-only} \verb+--divert-only+ 

Print diversions \emph{only}.  This forces OMake to be much more quiet,
printing output only when a command produces output.  With this option alone, the output from
successful commands is echoed immediately, and the output from failed commands is delayed until the
build completes and the summary is printed.  May be used in combination with \verb+--divert-repeat+
(output from failed commands will be printed twice).  Implies \verb+--divert-errors+.

You might also consider using the \verb+--progress+ flag (see Section~\ref{option:--progress}) so that you can see
when the build is active.

\option{--divert-discard-success} \verb+--divert-discard-success+

Discard the output from commands that were successful.  This can be useful in reducing
unwanted output so that you can concentrate on any errors.

\option{-d} \verb+-d [01jwWpPxXsS]+

For brevity, the \verb+-o+ option is also provided to duplicate the above output options.  The
\verb+-o+ option takes a argument consisting of a sequence of characters.  The characters are read
from left-to-right; each specifies a set of output options.  In general, an uppercase character turns
the option \emph{on}; a lowercase character turns the option \emph{off}.

\begin{description}
\item[0] Equivalent to \verb+-s --divert-only --divert-discard-success+

This option specifies that \verb+omake+ should be as quiet as possible.  If any errors occur
during the build, the output is delayed until the build terminates.  Output from successful commands
is discarded.

\item[1] Equivalent to \verb+-S --progress --divert-only --divert-repeat --divert-discard-success+

This is a slightly more relaxed version of ``quiet'' output.  The output from successful commands is
discarded.  The output from failed commands is printed immediately after the command complete.  The
output from failed commands is displayed twice: once immediately after the command completes, and
again when the build completes.  A progress bar is displayed so that you know when the build is
active.  Include the `\verb+p+' option if you want to turn off the progress bar (for example
\verb+omake -o 1p+).

\item[2] Equivalent to \verb+--progress --divert-only --divert-repeat+

The is even more relaxed, output from successful commands is printed.
This is often useful for deinterleaving the output when using \verb+-j+.

\item[W] Equivalent to \verb+-w+
\item[w] Equivalent to \verb+--no-w+
\item[P] Equivalent to \verb+--progress+
\item[p] Equivalent to \verb+--no-progress+
\item[X] Equivalent to \verb+--print-exit+
\item[x] Equivalent to \verb+--no-print-exit+
\item[S] Equivalent to \verb+-S+
\item[s] Equivalent to \verb+--no-S+
\end{description}

\section{Build options}

\option{-k} \verb+-k+

Do not abort when a build command fails;
continue to build as much of the project as possible.

\option{-n} \verb+-n+

This can be used to see what would happen if the project were to be built.

\option{-p} \verb+-p+

Watch the filesystem for changes, and continue the build until it succeeds.  If this
option is specified, \Prog{omake} will restart the build whenever source files are modified. Implies
\texttt{-k}.

\option{-P} \verb+-P+

Watch the filesystem for changes forever.  If this option is specified, \Prog{omake}
will restart the build whenever source files are modified. Implies
\texttt{-k}.

\option{-R} \verb+-R+

Ignore the current directory and build the project from its root directory.  When
\Prog{omake} is run in a subdirectory of a project, it normally builds files within the current
directory and its subdirectories.  If the \Opt{-R} option is specified, the build is performed as if
\Prog{omake} were run in the project root.

\option{-t} \verb+-t+

Update the \Prog{omake} database to force the project to be considered up-to-date.

\option{-U} \verb+-U+

Do not trust cached build information.  This will force the entire project to be rebuilt.

\option{--depend} \verb+--depend+

Do not trust cached dependency information.  This will force files to be rescanned
for dependency information.

\option{--configure} \verb+--configure+

Re-run \verb+static.+ sections of the included omake files, instead of
trusting the cached results.

\option{--force-dotomake} \verb+--force-dotomake+

Always use the \verb+$HOME/.omake+ for the \verb+.omc+ cache files.

\option{--dotomake} \verb+--dotomake <dir>+

Use the specified directory instead of the \verb+$HOME/.omake+
for the placement of the \verb+.omc+ cache files.

\option{-j} \verb+-j <count>+

Run multiple build commands in parallel.  The \Arg{count} specifies a
bound on the number of commands to run simultaneously.  In addition, the count may specify servers
for remote execution of commands in the form \verb+server=count+.  For example, the option
\verb+-j 2:small.host.org=1:large.host.org=4+ would specify that up to 2 jobs can be executed
locally, 1 on the server \verb+small.host.org+ and 4 on \verb+large.host.org+.  Each remote server
must use the same filesystem location for the project.

Remote execution is currently an experimental feature.  Remote filesystems like NFS do not provide
adequate file consistency for this to work.

\option{--print-dependencies} \verb+--print-dependencies+

Print dependency information for the targets on the command line.

\option{--show-dependencies} \verb+--show-dependencies <target>+

Print dependency information \emph{if} the \verb+target+ is built.

\option{--all-dependencies} \verb+--all-dependencies+

If either of the options \texttt{-{}-print-dependencies} or
\texttt{-{}-show-dependencies} is in effect, print transitive dependencies.  That is, print all
dependencies recursively.  If neither option \texttt{-{}-print-dependencies},
\texttt{-{}-show-dependencies} is specified, this option has no effect.

\option{--verbose-dependencies} \verb+--verbose-dependencies+

If either of the options \texttt{-{}-print-dependencies} or
\texttt{-{}-show-dependencies} is in effect, also print listings for each dependency.  The output is
very verbose, consider redirecting to a file.  If neither option \texttt{-{}-print-dependencies},
\texttt{-{}-show-dependencies} is specified, this option has no effect.

\option{--install} \verb+--install+

Install default files \File{OMakefile} and \File{OMakeroot} into the current directory.  You would
typically do this to start a project in the current directory.

\option{--install-all} \verb+--install-all+

In addition to installing files \File{OMakefile} and \File{OMakeroot}, install default
\File{OMakefile}s into each subdirectory of the current directory.  \Cmd{cvs}{1} rules are used for
filtering the subdirectory list.  For example, \File{OMakefile}s are not copied into directories
called \verb+CVS+, \verb+RCCS+, etc.

\option{--install-force} \verb+--install-force+

Normally, \Prog{omake} will prompt before it overwrites any existing \File{OMakefile}.  If this
option is given, all files are forcibly overwritten without prompting.

\option{--absname} \verb+--absname+

Filenames should expand to absolute pathnames.

\textbf{N.B.} This is an experimental option. It may become deprecated.

\option{variable definition} \verb+name=[value]+

\Prog{omake} variables can also be defined on the command line in the form \verb+name=value+.  For
example, the \verb+CFLAGS+ variable might be defined on the command line with the argument
\verb+CFLAGS="-Wall -g"+.

\section{Additional options}

In addition, \Prog{omake} supports a number of debugging flags on the command line. Run
\verb+omake --help+ to get a summary of these flags.

\section{Environment variables}

\subsection{\texttt{OMAKEFLAGS}}
\index{OMAKEFLAGS}

If defines, the \verb+OMAKEFLAGS+ should specify a set of options exactly as they are specified on
the command line.

\subsection{\texttt{OMAKELIB}}
\index{OMAKELIB}

If defined, the \verb+OMAKELIB+ environment variable should refer to the installed location of the
OMake standard library.  This is the directory that contains \verb+Pervasives.om+ etc.  On a Unix
system, this is often \verb+/usr/lib/omake+ or \verb+/usr/local/lib/omake+, and on Win32 systems it
is often \verb+c:\Program Files\OMake\lib+.

If not defined, \verb+omake+ uses the default configured location.  You should normally leave this
unset.

\section{Functions}

\subsection{\texttt{OMakeFlags}}

The \verb+OMakeFlags+ function can be used within an \verb+OMakefile+ to modify
the set of options.  The options should be specified exactly as they are on the command line.  For
example, if you want some specific project to be silent and display a progress bar, you can add the
following line to your \verb+OMakefile+.

\begin{verbatim}
    OMakeFlags(-S --progress)
\end{verbatim}

For options where it makes sense, the options are scoped like variables.  For example, if you want
OMake to be silent for a single rule (instead of for the entire project), you can use scoping the
restrict the range of the option.

\begin{verbatim}
    section
        # Do not display command output when foo is constructed
        OMakeFlags(-S)

        foo: fee
           echo "This is a generated file" > foo
           cat fee >> foo
           chmod 555 foo
\end{verbatim}

\section{Option processing}

When \verb+omake+ is invoked, the options are processed in the following order.

\begin{enumerate}
\item All options specified by the \verb+OMAKEFLAGS+ environment variable are defined globally.
\item All options from the command line are defined globally.
\item Any individual calls the the \verb+OMakeFlags+ function modify the options locally.
\end{enumerate}

\section{.omakerc}
\label{section:.omakerc}
\index{.omakerc}

If the \verb+$(HOME)/.omakerc+ exists, it is read before any of the \verb+OMakefiles+ in your
project.  The \verb+.omakerc+ file is frequently used for user-specific customization.
For example, instead of defining the \verb+OMAKEFLAGS+ environment variable, you could add
a line to your \verb+.omakerc+.

\begin{verbatim}
    $(HOME)/.omakerc:
        # My private options
        OMakeFlags(-S --progress)
\end{verbatim}    

% -*-
% Local Variables:
% Mode: LaTeX
% fill-column: 100
% TeX-master: "paper"
% TeX-command-default: "LaTeX/dvips Interactive"
% End:
% vim:tw=100:fo=tcq:
% -*-
